\sc Cells at War: The Playfulness of Game-Based Learning

### 2023-03-06 thesis_template/paper0904.tex
Mar 5 updates - fixed rosas mid-year report edits, started writing materials and methods

After many years of trying to adapt to educational innovations of the 21st Century, educators will soon be looking into the not too distant future and considering what modern learning in post-secondary education will look like in the 22nd Century. Technological advances will continue to transform the landscape of the classroom and the use of digital games, although a source of controversy and debate, will prove to be highly effective instructional tools that can promote problem solving skills and critical thinking. Their use has already started to change the structure of the class in the elementary classroom as they serve to energize and engage students in their learning and improve learning, creativity and imagination. The traditional construct of the classroom has started to evolve, embracing a new way in which knowledge, skills and attitudes can be acquired. Cells at War is a collection of scientific video games developed through interdisciplinary efforts that reinforces the notion that video games are highly effective tools used to enhance instructional methods. (Not only enhancing but teaching and engaging, ask rosa). – I don’t know if I should mention the game but I like this line.

@@

\subsection{Children and Play}

Although technological ecosystems have been established in schools for some time, its landscape continues to be missing significant components to create a fully functioning ecosystem. One of these key components is that educators are reticent to embrace digital technology and gaming as the new norm in the classroom. Kindergarten classrooms in Ontario are bustling places of activity, reflective of active inquiry through play. Exploration and provocations drive their curiosity while simultaneously developing their oral language and critical thinking skills \citep{kindergartencurriculum}. Digital games on ipads are being used to develop literacy and numeracy skills. Some educators however, oppose the use of digitized games especially in higher education and have not fully embraced the notion that learning can actually take place through play. We believe that play is paramount for children in their formative years and educational theorists like Vygotsky affirm this. He was firm in his belief that play is the leading line of development during preschool years \citep{vygotsky1967play}. Unfortunately, in many classrooms, educators continue to feel more comfortable in engaging in antiquated and traditional methods of instruction that consist of a didactic approach to learning. The need to challenge current educational practices is paramount as educators are currently teaching students from the perspective of their own past rather than the realities of the students’ future. – I like this line, its powerful that’s why I highlighted it.

@@

It can be argued that this fear of play is also due in large part to limited resources and training being provided to educators. Teachers require technological and pedagogical supports to develop their understanding of game based learning in the classroom. The expectation therefore to incorporate new and innovative teaching methodologies in the classroom is unrealistic without providing teachers with the appropriate training needed to do so. Incorporating game based learning in teacher practice is impacted by a lack of professional development opportunities that can systematically guide them in using games for teaching, learning, and assessment \citep{fishman2014empowering,ruggiero2013video,foster2020principles}. Might want to add a line here, don’t wanna end on quote

@@

The debate of whether game based learning helps or hinders learning is ongoing. What comes into question in opposition to game based learning is that it serves to act as a source of entertainment and amusement rather than for educative purposes. Many individuals associate play with children, and therefore believe that play is trivial and unimportant \citep{prensky2001fun}. In fact, one does not learn to play, but rather plays to learn. Play is where a spark is ignited and the source in which curiosity is peaked. It is not silly, trivial or irrelevant but has a biological and evolutionary function which has a strong connection to learning \citep{prensky2001fun}. It is, as Danny Hillis, founder of Thinking Machines and a former Disney Fellow has stated something every single culture does \citep{prensky2001fun}.

@@

Classrooms are becoming a place where the student is now involved in the learning process. Rules, goals and objectives, outcomes, competition, interaction, and story are what Mark Prensky refers to as the six key structural elements of games \citep{prensky2001fun}. These powerful factors are what set games apart from other engaging activities. Traditional forms of educational media lack what video games provide, a high degree of interactivity \citep{coller2009effectiveness}. Video games challenge individuals to work within specific parameters and respond to situations that occur in the game \citep{coller2009effectiveness}. The way in which individuals respond to challenges presented in video games allows them to think critically, strategize, and work collaboratively in an interactive and engaging manner.

@@

The growing evidence of the effectiveness of video games in the classroom is emerging, and the value of game based learning is evident throughout various disciplines. In 2005, Northern Illinois University began teaching an undergraduate course in Numerical Methods using a video game called “NIU-Torcs” \citep{coller2009effectiveness}. Students were observed spending twice the average amount of time on coursework than other subject areas \citep{coller2009effectiveness}. Other games are being developed for nursing education which focus on developing medication calculation skills. This is an area of global concern perhaps due to test anxiety and low mathematical self-efficacy in nursing students \citep{foss2013medication}.  Game based learning not only provides students with infinite possibilities for discovery but sustains student interest, builds confidence, and helps to improve cognition and learning outcomes.

\subsection{Concluding the Lit Review - I dont know if I wanna put this here}

The long-term positive effects of Game-Based Learning will continue to be researched and studied. Although empirical evidence is slowly emerging but still limited, studies and research to date are proving that Game-Based Learning has the potential to transform post secondary classrooms. They are moving from teacher centred to student centred learning hubs which serve to create a learning environment in which students are able to take risks, explore, discover, fail and try again.

NOW WE ADD CELLS AT WAR DATA!

>>>>

After many years of trying to adapt to educational innovations of the 21st Century, educators will soon be looking into the not too distant future and considering what modern learning in post-secondary education will look like in the 22nd Century. Technological advances will continue to transform the landscape of the classroom and the use of digital games, although a source of controversy and debate, will prove to be highly effective instructional tools that can promote problem solving skills and critical thinking. Their use has already started to change the structure of the class in the elementary classroom as they serve to energize and engage students in their learning and improve learning, creativity and imagination. The traditional construct of the classroom has started to evolve, embracing a new way in which knowledge, skills and attitudes can be acquired. Cells at War is a collection of scientific video games developed through interdisciplinary efforts that reinforces the notion that video games are highly effective tools used to enhance instructional methods and engage students in the Biology classroom.

@@

\subsection{Gamification of Learning}

Although the use of technology has become prevalent in schools across Canada, using game-based learning as an instructional approach is still in its infant stages. Educators are reticent to embrace digital technology and gaming as the new norm in the classroom due to insufficient technical competence, as well as challenges they face in selecting appropriate games \citep{jaaska2022teachers,molin2017role}. In addition, educators are rarely involved in the development of these games, making it difficult to facilitate suitable learning in the classroom \citep{molin2017role}.

Kindergarten classrooms in Ontario are bustling places of activity, reflective of active inquiry through play. Exploration and provocations drive their curiosity while simultaneously developing their oral language and critical thinking skills \citep{kindergartencurriculum}. We believe that play is paramount for children in their formative years and educational theorists like Vygotsky affirm this. He was firm in his belief that play is the leading line of development during preschool years \citep{vygotsky1967play}. Unfortunately, in many classrooms, educators continue to feel more comfortable engaging in antiquated and traditional methods of instruction that consist of a didactic approach to learning. Some educators may be fearful that incorporating game-based learning into the curriculum will undermine their authority in the classroom \citep{jaaska2022teachers,chee2014facilitating,jong2016teachers}. The need to challenge current educational practices is paramount as it appears some educators are currently teaching students from the perspective of their own past rather than the realities of the students’ future.

@@

It can be argued that this fear of play is also due in large part to limited resources and training being provided to educators \citep{fishman2014empowering}. Teachers require technological and pedagogical supports to develop their understanding of game based learning in the classroom. The expectation therefore to incorporate new and innovative teaching methodologies in the classroom is unrealistic without providing teachers with the appropriate training needed to do so. Incorporating game based learning in teacher practice is impacted by a lack of professional development opportunities that can systematically guide them in using games for teaching, learning, and assessment \citep{fishman2014empowering,ruggiero2013video,foster2020principles}. In order to develop teacher proficiency in game-based learning, further professional development is required in the form of pedagogical training in order to understand how to develop and utilize game-based learning practices.

@@

The debate of whether game based learning helps or hinders learning is ongoing. What comes into question in opposition to game based learning is that it serves to act as a source of entertainment and amusement rather than for educative purposes \citep{lister2015gamification, mayer2010adding}. Many individuals associate play with children, and therefore believe that play is trivial and unimportant \citep{prensky2001fun}. It appears one does not learn to play, but rather plays to learn. Play is where a spark is ignited and the source in which curiosity is peaked. It is not silly, trivial or irrelevant but has a biological and evolutionary function which has a strong connection to learning \citep{prensky2001fun}. It is, as Danny Hillis, founder of Thinking Machines and a former Disney Fellow has stated something every single culture does \citep{prensky2001fun}.

@@

 Rules, goals and objectives, outcomes, competition, interaction, and story are what Mark Prensky refers to as the six key structural elements of games \citep{prensky2001fun}. These powerful factors are what set games apart from other engaging activities. Traditional forms of educational media lack what video games provide, a high degree of interactivity \citep{coller2009effectiveness}. Video games challenge individuals to work within specific parameters and respond to situations that occur in the game \citep{coller2009effectiveness}. The way in which individuals respond to challenges presented in video games allows them to think critically, strategize, and work collaboratively in an interactive and engaging manner.

@@

The growing evidence of the effectiveness of video games in the classroom is emerging, and the value of game based learning is evident throughout various disciplines. In 2005, Northern Illinois University began teaching an undergraduate course in Numerical Methods using a video game called “NIU-Torcs” \citep{coller2009effectiveness}. Students were observed spending twice the average amount of time on coursework than other subject areas \citep{coller2009effectiveness}. Other games are being developed for nursing education which focus on developing medication calculation skills. This is an area of global concern perhaps due to test anxiety and low mathematical self-efficacy in nursing students \citep{foss2013medication}. Further, as a result of the COVID-19 pandemic, educators were compelled to find alternative ways to supplement hands-on learning experiences in an online setting. In a study completed at McMaster University, virtual reality (VR) headsets were used to engage in a 3D laboratory simulation which allowed for open-ended investigations in a range of STEM fields such as chemistry, biology, and physics \citep{tsirulnikov2023game}. \citet{tsirulnikov2023game} noted higher levels of engagement and motivation through these VR simulations. Game based learning not only provides students with infinite possibilities for discovery but sustains student interest, builds confidence, and helps to improve cognition and learning outcomes \citep{divjak2011impact}.

The long-term positive effects of Game-Based Learning will continue to be researched and studied. Although empirical evidence is slowly emerging but still limited, studies and research to date are proving that Game-Based Learning has the potential to transform post secondary classrooms. A learning environment in which students are able to take risks, explore, discover, fail and try again is achieved through a collaborative and interactive process between student and teacher.

\section{Materials and Methods}

\subsection{Study Setting and Context}

Results from this study were obtained utilizing a first-year cellular and molecular biology course (BIO1A03) at McMaster University (Ontario, Canada). BIO1A03 provides first year students with a strong cell and molecular biology foundation that is required for most upper level courses in the Faculty of Science at McMaster. Enrollment in BIO1A03 tends to be high, with approximately 700 students taking the course during both the fall (September-December) and winter (January-April) semesters. Due to the lengthy process of video game design, we were limited to collecting data solely from students enrolled in the winter semester for this study.

It is worthwhile to note the current course format of BIO1A03. Prior to 2014, BIO1A03 followed a traditional, didactic format, which consisted of three 50 minute in-person lectures per week, accompanied by a bi-weekly, three-hour laboratory session \citep{tahir2022blended}. In 2013, BIO1A03 was converted into a Blended Learning course format, which included independent, self-paced learning through prerecorded modules, and face-to-face learning in the classroom with a heavier focus on student collaboration and engagement \citep{tahir2022blended}. Course content previously delivered during lecture was condensed into two weekly modules which students were required to complete prior to attending in-person lectures. In-person lectures were restructured and minimized to two per week. One of these sessions consisted of a review lecture, where students could clarify concepts they found difficult to understand, while the other focused on applying theoretical concepts that students had learned to a real world concept. In a study highlighting the impacts of blended learning in BIO1A03 at McMaster, \citet{tahir2022blended} noted that the majority of students surveyed after taking the course felt that these novel blended learning approaches enabled them to improve general knowledge and understanding of core concepts in cellular and molecular biology.

\subsection{Video Game Design/Creation}

The conversion to a blended learning approach in BIO1A03 allowed for a seamless transition to online-learning in 2019 when the COVID-19 pandemic began. During this time, the BIO1A03 instructional team began exploring other novel pedagogical approaches which could help improve student learning. The idea of using digital games to improve education has been around for quite some time, however the controversy surrounding video games in this context is still apparent.

In 2021, the BIO1A03 instructional team partnered up with the George Brown College School of Design, and received support from CEWIL Canada (Co-operative Education and Work-Integrated Learning Canada), formerly known as CAFCE (Canadian Association for Co-operative Education), to create and develop an educational biological video game that could be used in BIO1A03.

\subsection{Study Participants}

\subsection{Survey Design: Qualitative Analysis of Student Feedback}

\section{Results}

\section{Discussion}

\section{Future Directions}
